---
title: "golub"
author: "Libby Reeves"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction to the Golub data set

TODO clean this up
Gene expression data (3051 genes and 38 tumor mRNA samples) from the leukemia microarray study of Golub et al. (1999). Pre-processing was done as described in Dudoit et al. (2002). Some patients had ALL (acute lymphoblastic leukemia) while the rest had AML (acute myeloid leukemia).

## Loading the Golub data set and formatting for use
```{r warning=FALSE, message=FALSE}
library(golubEsets)
data(Golub_Train)

data <- exprs(Golub_Train)
ann <- pData(Golub_Train)
```

Dimensions and data from the golub dataset:
```{r}
dim(data)
data[1:5][1:5]
```

Length and content of the golub annotation (classes):
```{r}
dim(ann)
colnames(ann)
ann[1:5][1:5]
```

## Hierarchical clustering and correlation of samples with heatmap

```{r fig.height=8, fig.width=8}
colnames(data) <- paste(ann$Samples, ann$ALL.AML, sep='-')
data.cor <- cor(data)
heatmap(data.cor, cexRow=0.7, cexCol=0.7)

dist.samples <- dist(t(data))
hc.samples <- hclust(dist.samples, method='median')
plot(hc.samples, cex=0.8)

```

## PCA
```{r}
data.pca <- prcomp(t(data))
data.loadings <- data.pca$x

data.pca.var <- round(data.pca$sdev^2 / sum(data.pca$sdev^2)*100, 2)
plot(c(1:length(data.pca.var)), data.pca.var, type="b", xlab="# components",
     ylab="% variance", pch=21, col=1, bg=3, cex=1.5,
     main="Scree plot showing % variability explained by each eigenvalue")
```
```{r fig.height=10, fig.width=7}
par(mfrow=c(3, 1))

plot(data.loadings[, 1], data.loadings[, 2], xlab='PC1', ylab='PC2', pch=19,
     col=as.numeric(as.factor(ann$ALL.AML)), main="PC2 vs. PC1")
legend('topleft', legend=levels(as.factor(ann$ALL.AML)),
       col=1:length(levels(as.factor(ann$ALL.AML))), pch=c(19, 19))

plot(data.loadings[, 1], data.loadings[, 3], xlab='PC1', ylab='PC3', pch=19,
     col=as.numeric(as.factor(ann$ALL.AML)), main="PC3 vs. PC1")
legend('topleft', legend=levels(as.factor(ann$ALL.AML)),
       col=1:length(levels(as.factor(ann$ALL.AML))), pch=c(19, 19))

plot(data.loadings[, 2], data.loadings[, 3], xlab='PC2', ylab='PC3', pch=19,
     col=as.numeric(as.factor(ann$ALL.AML)), main="PC3 vs. PC2")
legend('topleft', legend=levels(as.factor(ann$ALL.AML)),
       col=1:length(levels(as.factor(ann$ALL.AML))), pch=c(19, 19))

```

## PCA within ALL data: B-cell vs. T-cell

```{r}
data.all <- data[, ann$ALL.AML=='ALL']

data.all.pca <- prcomp(t(data.all))
data.all.loadings <- data.all.pca$x

data.all.pca.var <- round(data.all.pca$sdev^2 / sum(data.all.pca$sdev^2)*100, 2)
plot(c(1:length(data.all.pca.var)), data.all.pca.var, type="b", 
     xlab="# components", ylab="% variance", pch=21, col=1, bg=3, cex=1.5,
     main="Scree plot showing % variability explained by each eigenvalue")

```


```{r fig.height=10, fig.width=7}
par(mfrow=c(3, 1))
t.b.cell <- as.factor(ann$T.B.cell)

plot(data.all.loadings[, 1], data.all.loadings[, 2], xlab='PC1', ylab='PC2',
     pch=19, col=as.numeric(t.b.cell), main="PC2 vs. PC1")
legend('bottomright', legend=levels(t.b.cell),
       col=1:length(levels(t.b.cell)), pch=c(19, 19))

plot(data.all.loadings[, 1], data.all.loadings[, 3], xlab='PC1', ylab='PC3',
     pch=19, col=as.numeric(t.b.cell), main="PC3 vs. PC1")
legend('bottomright', legend=levels(t.b.cell),
       col=1:length(levels(t.b.cell)), pch=c(19, 19))

plot(data.all.loadings[, 2], data.all.loadings[, 3], xlab='PC2', ylab='PC3',
     pch=19, col=as.numeric(t.b.cell), main="PC3 vs. PC2")
legend('bottomright', legend=levels(t.b.cell),
       col=1:length(levels(t.b.cell)), pch=c(19, 19))

```
