---
title: "golub"
author: "Libby Reeves"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction to the Golub data set

TODO clean this up
Gene expression data (3051 genes and 38 tumor mRNA samples) from the leukemia microarray study of Golub et al. (1999). Pre-processing was done as described in Dudoit et al. (2002). Some patients had ALL (acute lymphoblastic leukemia) while the rest had AML (acute myeloid leukemia).

## Loading the Golub data set and formatting for use
```{r warning=FALSE, message=FALSE}
library(golubEsets)
data(Golub_Train)

data <- exprs(Golub_Train)
ann <- pData(Golub_Train)
```

Dimensions and data from the golub dataset:
```{r}
dim(data)
data[1:5][1:5]
```

Length and content of the golub annotation (classes):
```{r}
dim(ann)
colnames(ann)
ann[1:5][1:5]
```

## Hierarchical clustering and correlation of samples with heatmap

```{r fig.height=8, fig.width=8}
colnames(data) <- paste(ann$Samples, ann$ALL.AML, sep='-')
data.cor <- cor(data)
heatmap(data.cor, cexRow=0.7, cexCol=0.7)

dist.samples <- dist(t(data))
hc.samples <- hclust(dist.samples, method='median')
plot(hc.samples, cex=0.8)

```

## PCA
```{r}
data.pca <- prcomp(t(data))
data.loadings <- data.pca$x

data.pca.var <- round(data.pca$sdev^2 / sum(data.pca$sdev^2)*100, 2)
plot(c(1:length(data.pca.var)), data.pca.var, type="b", xlab="# components",
     ylab="% variance", pch=21, col=1, bg=3, cex=1.5,
     main="Scree plot showing % variability explained by each eigenvalue")
```
```{r fig.height=10, fig.width=7}
par(mfrow=c(3, 1))

plot(data.loadings[, 1], data.loadings[, 2], xlab='PC1', ylab='PC2', pch=19,
     col=as.numeric(as.factor(ann$ALL.AML)), main="PC2 vs. PC1")
legend('topleft', legend=levels(as.factor(ann$ALL.AML)),
       col=1:length(levels(as.factor(ann$ALL.AML))), pch=c(19, 19))

plot(data.loadings[, 1], data.loadings[, 3], xlab='PC1', ylab='PC3', pch=19,
     col=as.numeric(as.factor(ann$ALL.AML)), main="PC3 vs. PC1")
legend('topleft', legend=levels(as.factor(ann$ALL.AML)),
       col=1:length(levels(as.factor(ann$ALL.AML))), pch=c(19, 19))

plot(data.loadings[, 2], data.loadings[, 3], xlab='PC2', ylab='PC3', pch=19,
     col=as.numeric(as.factor(ann$ALL.AML)), main="PC3 vs. PC2")
legend('topleft', legend=levels(as.factor(ann$ALL.AML)),
       col=1:length(levels(as.factor(ann$ALL.AML))), pch=c(19, 19))

```

## PCA within ALL data: B-cell vs. T-cell

```{r}
data.all <- data[, ann$ALL.AML=='ALL']

data.all.pca <- prcomp(t(data.all))
data.all.loadings <- data.all.pca$x

data.all.pca.var <- round(data.all.pca$sdev^2 / sum(data.all.pca$sdev^2)*100, 2)
plot(c(1:length(data.all.pca.var)), data.all.pca.var, type="b", 
     xlab="# components", ylab="% variance", pch=21, col=1, bg=3, cex=1.5,
     main="Scree plot showing % variability explained by each eigenvalue")

```


```{r fig.height=10, fig.width=7}
par(mfrow=c(3, 1))
t.b.cell <- as.factor(ann$T.B.cell)

plot(data.all.loadings[, 1], data.all.loadings[, 2], xlab='PC1', ylab='PC2',
     pch=19, col=as.numeric(t.b.cell), main="PC2 vs. PC1")
legend('bottomright', legend=levels(t.b.cell),
       col=1:length(levels(t.b.cell)), pch=c(19, 19))

plot(data.all.loadings[, 1], data.all.loadings[, 3], xlab='PC1', ylab='PC3',
     pch=19, col=as.numeric(t.b.cell), main="PC3 vs. PC1")
legend('bottomright', legend=levels(t.b.cell),
       col=1:length(levels(t.b.cell)), pch=c(19, 19))

plot(data.all.loadings[, 2], data.all.loadings[, 3], xlab='PC2', ylab='PC3',
     pch=19, col=as.numeric(t.b.cell), main="PC3 vs. PC2")
legend('bottomright', legend=levels(t.b.cell),
       col=1:length(levels(t.b.cell)), pch=c(19, 19))

```

## Testing for differentially expressed genes

```{r}
wilcox.test.all.genes <- function(x, s1, s2) {
	x1 <- as.numeric(x[s1])
	x2 <- as.numeric(x[s2])
	w.out <- wilcox.test(x1, x2, exact=F, alternative='two.sided', correct=T)
	#out <- as.numeric(w.out$statistic)
	out <- as.numeric(w.out$p.value)
	#print(p.val)
	return(out)
}

wilcox.results <- apply(data, 1, wilcox.test.all.genes,
                    s1 = ann$ALL.AML=='ALL', s2 = ann$ALL.AML=='AML')

head(wilcox.results)
```

```{r}
# multiple testing correction

# Benjamini-Hochberg
wilcox.bh <- p.adjust(wilcox.results, method='BH')

wilcox.bh.sig.05 <- wilcox.bh[wilcox.bh < 0.05]
wilcox.bh.sig.01 <- wilcox.bh[wilcox.bh < 0.01]

length(wilcox.bh.sig.01)

# Bonferroni - more conservative
wilcox.bonferroni <- p.adjust(wilcox.results, method='bonferroni')
wilcox.bonferroni.sig.05 <- wilcox.bonferroni[wilcox.bonferroni < 0.05]

length(wilcox.bonferroni.sig.05)
```

## Some genes of interest in AML and ALL

```{r}
library(hu6800.db)
library(AnnotationDbi)

genes.of.interest <- c('FLT3', 'NPM1', 'IDH1', 'IDH2', 'DNMT3A', 'TET2',
                       'ASXL1', 'TP53', 'NOTCH1', 'PAX5', 'IKZF1')

probe.ids <- featureNames(Golub_Train)
gene.symbols <- mapIds(
  x = hu6800.db,           # The chip database
  keys = probe.ids,        # Your array probe IDs
  column = "SYMBOL",       # Target identifier type
  keytype = "PROBEID",     # Source identifier type
  multiVals = "first"      # How to handle multi-mapping probes
)
gene.symbols <- na.omit(gene.symbols)
symbols.of.interest <- gene.symbols[gene.symbols %in% genes.of.interest]
symbols.of.interest
```

## Genes with most significant expression difference between ALL and AML

```{r}
sig.probe.ids <- names(wilcox.bonferroni.sig.05)

sig.genes <- gene.symbols[names(gene.symbols) %in% sig.probe.ids]

sig.data <- data.frame('probe_id'=names(sig.genes), 'gene_name'=sig.genes,
                       'p_value'=wilcox.bonferroni.sig.05)
sig.data
# TODO maybe remove row names?
```

## Gene expression box plots

```{r fig.width=12, fig.height=10}

par(mfrow=c(4, 3))
make.boxplots <- function(gene.symbol) {
  probe.id <- names(gene.symbols[gene.symbols==gene.symbol])[1]
  gene.data <- data[probe.id, ]
  length(gene.data)
  boxplot(gene.data ~ ann$ALL.AML, col=c('red', 'blue'), xlab='',
          main=paste(gene.symbol, 'expression'))
}

for (symbol in sig.data$gene_name) {
  make.boxplots(symbol)
}


```






